Practical Optional Types for Clojure

### 2015-02-25 overview.tex
finish overview

\citet{TF10}
introduced Typed Racket with occurrence typing,
an inference technique for conditional control flow.
They introduce the concept of occurrence typing 
with the following example.

@@

To check this example, we note that
occurrence typing already provides
valuable information about the conditional expression---namely that it cannot return logically true and if it returns logically false 
then \clj{x} is a \clj{Number}. Equivalently, if it returns a value then \clj{x} is a \clj{Number}.
We use this assumption to check \clj{(inc x)}, automatically
ruling out the possibility of null-pointer exceptions
(said formally in section~\ref{sec:doformal}).

@@

\begin{Code}
\begin{exmp}
\inputminted[firstline=6,lastline=8]{clojure}{code/demo/src/demo/hmap.clj}
\inputminted[firstline=26,lastline=28]{clojure}{code/demo/src/demo/hmap.clj}
\end{exmp}
\end{Code}

@@

HMaps support various options to track the presence and absence of entries.
\clj{'{:a Num}} is a shorthand for the \clj{:mandatory} option---entries that
must be present.

\begin{Code}
\begin{exmp}
\inputminted[firstline=58,lastline=64]{clojure}{code/demo/src/demo/hmap.clj}
\end{exmp}
\end{Code}

Looking up \clj{:a} simply returns the associated type.
Keys other than \clj{:a} are unaccounted for in \clj{Mand}, so 
the best we can infer is \clj{Any}, the supertype of all types.

\clj{:absent-keys} is a set of keywords that are known to not have entries.
We take advantage of this to infer a better type for lookups on known absent keys.

\begin{Code}
\begin{exmp}
\inputminted[firstline=66,lastline=73]{clojure}{code/demo/src/demo/hmap.clj}
\end{exmp}
\end{Code}

\clj{:optional} is a map that declares entries either absent or
present with a specific type.

@@

\clj{:complete?} is \clj{true} if there are no further entries
than those declared \clj{:mandatory} or \clj{:optional}.

\begin{Code}
\begin{exmp}
\inputminted[firstline=17,lastline=23]{clojure}{code/demo/src/demo/hmap_path.clj}
\end{exmp}
\end{Code}

@@

\clj{enc-keypair} takes an unencrypted keypair and returns an encrypted keypair by
removing the raw private key and attaching an encrypted private key.
Since \clj{EncKeyPair} is \clj{:complete?}, Typed Clojure enforces the return type
does not contain an entry \clj{:private-key}, and would complain if the \clj{dissoc}
operation forgot to remove it.

@@

In the then branch Typed Clojure infers \clj{'{:a Num}}, in the else branch
\clj{(HMap :optional {:a nil})}. Failed lookups return \clj{nil},
so we cannot distinguish between \clj{'{:a nil}} and \clj{(HMap :optional {:a nil})}
by testing the result of looking up \clj{:a}.
Section~\ref{sec:hmapformal} discusses this in greater detail.

>>>>

\citet{TF10}
presented Typed Racket with occurrence typing,
an inference technique for conditional control flow.
They introduce the concept of occurrence typing 
with the following example.

@@

To check this example, we note that
occurrence typing already provides
valuable information about the conditional expression---namely that it cannot return logically true and if it returns logically false 
then \clj{x} is a \clj{Number}. Equivalently, if it returns a value then \clj{x} is a \clj{Number}.
We use this assumption to check \clj{(inc x)} with an automatic
guarantee that a null-pointer exception is impossible
(modelled formally in section~\ref{sec:doformal} and proved
in section~\ref{sec:metatheory}).

@@

\begin{Code}
\begin{exmp}
\inputminted[firstline=6,lastline=8]{clojure}{code/demo/src/demo/hmap.clj}
\inputminted[firstline=27,lastline=29]{clojure}{code/demo/src/demo/hmap.clj}
\end{exmp}
\end{Code}

@@

HMaps support various options to track the presence and absence of entries.
\clj{:mandatory} declares entries that must be present.

\begin{Code}
\begin{exmp}
\inputminted[firstline=58,lastline=64]{clojure}{code/demo/src/demo/hmap.clj}
\end{exmp}
\end{Code}
Looking up \clj{:a} simply returns the associated type \clj{Num}.
Keys other than \clj{:a} are unaccounted for in \clj{Mand}, so 
the best we can infer is \clj{Any}, the supertype of all types.
\clj{'{:a Num}} is shorthand for \clj{(HMap :mandatory {:a Num})}.

\clj{:absent-keys} is a set of keywords that are known to not have entries.
\begin{Code}
\begin{exmp}
\inputminted[firstline=66,lastline=73]{clojure}{code/demo/src/demo/hmap.clj}
\end{exmp}
\end{Code}
We can infer a better type for \clj{(:b m)}---\clj{:b} is absent in \clj{Abs}
so it looking up \clj{:b} returns \clj{nil}.

\clj{:optional} is a map that declares entries either absent \emph{or}
present with a specific type.

@@

\clj{:complete?} is \clj{true} if there are no further entries
than those declared \clj{:mandatory} or \clj{:optional}.
\begin{Code}
\begin{exmp}
\inputminted[firstline=17,lastline=26]{clojure}{code/demo/src/demo/hmap_path.clj}
\end{exmp}
\end{Code}
\clj{:a} is \clj{:mandatory} in \clj{Comp} so lookups return \clj{Num}.
\clj{:b} is \clj{:optional} so lookups return \clj{(U nil Kw)}.
\clj{:c} is not mentioned, but since \clj{:complete?} is \clj{true}
it must be absent so a lookup returns \clj{nil}.

@@

\clj{enc-keypair} takes an unencrypted keypair and returns an encrypted keypair by
removing the raw \clj{:private-key} entry and associating an encrypted private key
as \clj{:enc-private-key}.
Since \clj{EncKeyPair} is \clj{:complete?}, Typed Clojure enforces the return type
does not contain an entry \clj{:private-key}, and would complain if the \clj{dissoc}
operation forgot to remove it.

@@

Typed Clojure infers \clj{m} as \clj{'{:a Num}} in the then branch
and \clj{(HMap :optional {:a nil})}
in the else branch.
We cannot infer 
\begin{minted}{clojure}
 (HMap :absent-keys #{:a})
\end{minted}
in the else branch because \clj{m} could be the value \clj{{:a nil}}
by the definiton of \clj{Opt}.
Section~\ref{sec:hmapformal} discusses this in greater detail.
